Cost- and Deadline-Constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-04-19 paper/paper.tex
Add text from Gideon's thesis

\usepackage[pdftex]{graphicx}
\graphicspath{{./figures/}}
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\usepackage[cmex10]{amsmath}
\usepackage{algorithm,algpseudocode}
\usepackage{hyperref} % must come before url
\usepackage{url}

@@

Scientific workflows, usually represented as directed acyclic graphs (DAGs), are
an important class of distributed applications that have been studied in the context of
resource management and task scheduling on grid and utility computing systems.
Workflow applications for large computational problems are often composed of 
several inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ in their
input data, number of tasks, and individual task sizes. The CyberShake 
application~\cite{Callaghan11} is a good example of an application that requires 
scientific workflow ensembles. CyberShake computes seismic hazard curves for a given
geographical location. In order to produce a hazard map for a region such as 
Southern California, many hazard curves need to be computed for a large number of 
geographical sites, where each site requires running a single, complex scientific 
workflow. Workflows for individual sites may differ not only in their parameters, 
but also in their priority. For example, some sites may be in heavily populated areas 
or in strategic locations such as power plants, while others may be less important. 
Other examples of workflow ensembles come from astronomical applications that create
science-grade image mosaics~\cite{Deelman08} or search for extrasolar 
planets~\cite{vockler11}, where multiple, different workflows are used to
process data for different areas of the sky.

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost, on-demand computing resources~\cite{Ostermann10,Keahey09}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give users more flexibility in creating a controlled and managed computing environment.
Clouds provide the ability to adjust resource capacity according to the dynamically changing
computing demands of the application, often called auto-scaling. However, giving users 
more control over the resources also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as workflow makespan or resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) cloud providers 
usually have monetary costs associated with them~\cite{Durkee10}.

In this paper, we aim to gain insight into resource management challenges when executing 
scientific workflow ensembles on clouds. In particular, we address a new and important 
problem of maximizing the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation for this work is to answer the fundamental question of 
concern to a researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble based on the constraints. We evaluate these algorithms using a
simulator we have developed using CloudSim~\cite{Calheiros11}, which models the 
infrastructure and the application. The simulator includes also tightly-coupled 
scheduler and provisioner modules. The algorithms were evaluated on a set 
of scientific workflow ensembles from several different applications using a 
broad range of budget and deadline parameters.

@@

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan10,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in the scenarios we consider in this paper. Our work can be also
regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In
this model, virtual machine (VM) instances are requested on-demand. The VMs are
billed by the hour, with partial hours being rounded up. Although there are
multiple VM types with different amounts of CPU, memory, disk space, and I/O,
for this paper we focus on a single VM type because we assume that for most
applications there will typically be one VM type with the best price/performance
ratio for the application~\cite{Juve2009}. We also assume that a submitted task
has exclusive access to a VM instance and that there is no preemption. For this
study we assumed that the delays in VM start and terminate requests are negligible.

\subsection{Application Model}
The target applications for this paper are scientific workflows that can be
modeled as Directed Acyclic Graphs (DAGs), where the nodes in the graph
represent computational tasks, and the edges represent data- or control-flow
dependencies between the tasks.

We assume that we have runtime estimates for each task in the workflow based on
either a performance model of the application, or historical data that can be
mined. In this study we use synthetic workflows that were generated using 
historical data from real applications~\cite{Bharathi08}. These applications 
come from a wide variety of domains including: bioinformatics (Epigenomics, 
SIPHT: sRNA identification protocol using high-throughput
technology), astronomy (Montage), earthquake science (CyberShake), and physics 
(LIGO).

Although scientific workflows are often data-intensive, we do not currently
consider the size of input and output data when scheduling tasks. Instead we
assume that all workflow data is stored in a shared cloud storage system, such
as Amazon S3, and that data transfer times are included in task runtime
estimates. We also assume that data transfer times between the shared storage
and the VMs are equal for different VMs so that task placement decisions do not
have an impact on the runtime of the tasks.

This paper focuses on scheduling and provisioning for workflow ensembles. A
workflow ensemble consists of several related workflows that a user wishes to
run. Each workflow in an ensemble is given a priority that indicates how
important the workflow is to the user. These priorities are absolute in the 
sense that completing a high priority workflow is more valuable than 
completing all lower priority workflows combined.

The goal of the workflow ensemble scheduling and cloud provisioning problem
considered in this paper is to complete as many high-priority workflows as
possible given a fixed budget and deadline. Only workflows for which all tasks
are finished by the deadline are considered to be complete---partial results are
not usable in this model.

@@

In this section we describe several algorithms we developed to execute ensembles of
workflows on the cloud under budget and deadline constraints.

\subsection{Static Provisioning Dynamic Scheduling \\
(SPDS) }
\label{sec:spds}

The simplest strategy for executing ensembles of workflows in the cloud is to
provision resources statically and schedule workflow tasks dynamically. Given a 
budget $b$, deadline $d$, and the hourly price $p$ of a VM, it is easy to calculate
the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before
the deadline:

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}

The SPDS algorithm statically provisions $N_{VM}$ VMs at the start of the ensemble execution 
and keeps them running until the deadline is reached, or the budget runs out.
This provisioning plan has the advantage that it minimizes the number of provisioning 
and de-provisioning requests.

@@

Initially, the ready tasks from all workflows in the ensemble are added to a 
priority queue based on the priority of the workflow to which they belong. If
there are idle VMs available, and the priority queue is not empty, the next task
from the priority queue is submitted to a randomly chosen idle VM. The process is 
repeated until there are no idle VMs or the priority queue is empty. The scheduler
then waits for a task to finish, adds its ready children to the priority queue,
marks the VM as idle, and the entire process repeats until the deadline is reached.

@@

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{spds-gantt}
 \caption{Example schedule generated using SPDS algorithm. The colors indicate how various tasks from different workflows are scheduled onto Virtual Machines. }
\label{fig:spds-example}
\end{figure}

\subsection{Dynamic Provisioning Dynamic \\Scheduling (DPDS)}

The static provisioning approach of SPDS may perform poorly when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing 
is that the number of resources can be changed to meet the demands of an application.
In policy-based provisioning (or auto-scaling) systems, metrics such as resource 
utilization or task queue length are monitored to estimate application demand.
The DPDS algorithm uses this approach to determine when to provision and de-provision VMs. The 
algorithm periodically computes resource utilization as the percentage of idle VMs, and
adjusts the number of VMs if the utilization is above or below given thresholds.
Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate. This process
is described in Algorithm~\ref{alg:prov}.

@@

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated in line \ref{l:terminate2} are chosen to increase the resource
utilization to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the number of running VMs is greater than the product of
$N_{VM}$ and an autoscaling factor, $v_{max}$ that has to be set manually.

@@

The DPDS algorithm does not use any information about the structure of the workflows
in the ensemble when scheduling tasks: it looks only at the priorities of the ready 
tasks when deciding which task to schedule next. It does not consider whether a lower 
priority task belongs to a workflow that will never be able to complete given the 
current budget and deadline. As a result, DPDS may execute lower priority tasks just 
to keep VMs busy and that will end up delaying higher priority tasks, making it less likely
that higher priority workflows will be able to finish.

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends 
DPDS by introducing a workflow admission procedure. The admission procedure is
invoked whenever WA-DPDS sees the first task of a new workflow at the head of the 
priority queue (i.e. when no other tasks from the workflow have been scheduled yet). 
The admission procedure -- shown in Algorithm~\ref{alg:wa-dpds} -- estimates
whether there is enough budget remaining to admit the new workflow; if there is not, then the 
workflow is rejected and its tasks are removed from the queue. This algorithm 
compares the current cost (consumed budget) and remaining budget, taking
into account the cost of currently running VMs, and the cost of workflows 
that have already been admitted. It adds a small safety margin to  avoid going over
the budget.

@@

\medskip
\medskip

\subsection{Static Provisioning Static Scheduling \\(SPSS)}

@@

The approach used by SPSS is to plan each workflow in the ensemble in priority 
order, rejecting any workflows that cannot be completed by the deadline or that 
cause the cost of the plan to exceed the budget. Once the plan is complete, 
the VMs are provisioned and tasks are executed according to the schedules given
by the plan. However, in this paper we only consider a deterministic environment and task times.

The disadvantage of the static planning approach used by SPSS is that it is 
sensitive to dynamic changes in the environment and the application that may 
disrupt the carefully constructed plan.

@@

SPSS assigns sub-deadlines to each task based on the slack time of the
workflow, which is defined as the amount of extra time that a workflow
can extend its critical path and still be completed by the ensemble deadline.
For a workflow $w$, the slack time of $w$ is:

$$
ST(w) = d - CP(w)
$$

where $d$ is the ensemble deadline and $CP(w)$ is the critical path of $w$.
We assume that $CP(w) \leq d$, otherwise the workflow cannot be completed 
by the deadline and must be rejected. For large ensembles we expect the 
critical path of any individual workflow to be much less than the deadline.

A task's level is the length of the longest path between the task and an entry
task of the workflow:

$$
Level(t) = \begin{cases}
0,&$if~$Pred(t) = \emptyset\cr
\max_{p \in Pred(t)}Level(p)+1,&$otherwise.$
\end{cases}
$$

SPSS distributes the slack time of the workflow by level, so that each level
of the workflow gets a portion of the workflow's slack time proportional to
the number of tasks in the level and the total runtime of tasks in the level.
The idea is that levels containing many tasks and large runtimes should be given
a larger portion of the slack time so that tasks in those levels may be serialized.
Otherwise, many resources need to be allocated to run all of the tasks in 
parallel, which may be more costly.

The slack time of a level $l$ in workflow $w$ is given by:

$$
ST(l) = ST(w) \Biggl[\left({\alpha}\frac{N(l)}{N(w)}\right) + \left({(1 - \alpha)}\frac{R(l)}{R(w)} \right)\Biggr]
$$

where $N(w)$ is the number of tasks in the workflow, $N(l)$ is the number of tasks 
in level $l$, $R(w)$ is the total runtime of all tasks in the workflow, $R(w)$ is
the total runtime of all tasks in level $l$, and $\alpha$ is a parameter between
0 and 1 that causes more slack time to be given to levels with more tasks (large 
$\alpha$) or more runtime (small $\alpha$).

The deadline of a task $t$ is then:

$$
DL(t) = LST(t) + RT(t) + ST(Level(t))
$$

where $Level(t)$ is the level of $t$, $RT(t)$ is the runtime of $t$, and $LST(t)$ 
is the latest start time of $t$ determined by:

$$
LST(t) = \begin{cases}
0,&$if~$Pred(t) = \emptyset\cr
\max_{p \in Pred(t)} DL(p),&$otherwise.$
\end{cases}
$$

@@

\begin{figure}[tb] 
\centering
\includegraphics[width=1.0\columnwidth]{spss-gantt}
 \caption{Example of schedule generated using SPSS algorithm. Tasks are colored by workflow. }
\label{fig:spss-example}
\end{figure}

\section{Performance Evaluation}
\label{sec:performance}

@@

To evaluate and compare our algorithms, we developed a cloud workflow simulator based on
CloudSim~\cite{Calheiros11}. Our simulation model consists of Cloud, VM and WorkflowEngine 
entities. The Cloud entity is responsible for staring and terminating VM entities using 
an API similar to Amazon EC2. VM entities simulate the execution of individual tasks,
and the WorkflowEngine entity manages the scheduling of tasks and the provisioning of
VMs. We assume that the VMs have a single core and execute tasks sequentially. Although 
CloudSim provides a more advanced infrastructure model which includes time-sharing and 
space-sharing policies, and power consumption in a datacenter, we do not use these 
features, since we are interested mainly in the execution of tasks on VMs and high-level 
workflow scheduling and provisioning. The simulator reads workflows in a modified version of
the DAX format used by the Pegasus Workflow Management System~\cite{Deelman2005}. The modified format was
created for the synthetic workflow generator developed by Bharathi, et al~\cite{Bharathi08}.

@@

\begin{figure}[tb] 
\centering
\includegraphics[width=0.6\columnwidth]{ensemble-pareto}
\caption{Pareto-like distribution of workflow sizes in ensemble. Size is
measured in number of tasks.}
\label{fig:ensemble-distribution}
\end{figure}

@@

In order to evaluate the algorithms on a standard set of workflows, we used
ensembles constructed from the workflows available\footnote{\url{https://confluence.pegasus.isi.edu/display/pegasus/WorkflowGenerator}}
in the workflow generator gallery~\cite{Bharathi08}. The gallery contains synthetic
workflows modeled on real applications. We used workflows from five applications 
that were managed by the Pegasus Workflow Management System. For each
application we constructed ensembles of 100 workflows by sampling collections of
workflows of various sizes (number of tasks) based on a given distribution.
Two distributions are considered:

@@

>>>>

\usepackage[pdftex]{graphicx}
\graphicspath{{./figures/}}
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\usepackage[cmex10]{amsmath}
\usepackage{algorithm,algpseudocode}
\usepackage{hyperref} % must come before url
\usepackage{url}
\usepackage{color} % for \textcolor
\usepackage{subfig} % for \subfloat

\newcommand{\TODO}[1]{
  {\Large \textcolor{red}{\textbf{TODO: }#1}}
}

@@

Scientific workflows, usually represented as directed acyclic graphs (DAGs), are
an important class of distributed applications that have been studied in the context of
resource management and task scheduling on grid and utility computing systems.
Workflow applications for large computational problems are often composed of 
several inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ in their
input data, number of tasks, and individual task sizes. The CyberShake 
application~\cite{Callaghan2011} is a good example of an application that requires 
scientific workflow ensembles. CyberShake computes seismic hazard curves for a given
geographical location. In order to produce a hazard map for a region such as 
Southern California, many hazard curves need to be computed for a large number of 
geographical sites, where each site requires running a single, complex scientific 
workflow. Workflows for individual sites may differ not only in their parameters, 
but also in their priority. For example, some sites may be in heavily populated areas 
or in strategic locations such as power plants, while others may be less important. 
Other examples of workflow ensembles come from astronomical applications that create
science-grade image mosaics~\cite{Deelman2008} or search for extrasolar 
planets~\cite{Vockler2011}, where multiple, different workflows are used to
process data for different areas of the sky.

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost, on-demand computing resources~\cite{Ostermann2010, Keahey2009}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give users more flexibility in creating a controlled and managed computing environment.
Clouds provide the ability to adjust resource capacity according to the dynamically changing
computing demands of the application, often called auto-scaling. However, giving users 
more control over the resources also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as workflow makespan or resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) cloud providers 
usually have monetary costs associated with them~\cite{Durkee2010}.

In this paper, we aim to gain insight into resource management challenges when executing 
scientific workflow ensembles on clouds. In particular, we address a new and important 
problem of maximizing the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation for this work is to answer the fundamental question of 
concern to a researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble based on the constraints. We evaluate these algorithms using a
simulator we have developed using CloudSim~\cite{Calheiros2011}, which models the 
infrastructure and the application. The simulator includes also tightly-coupled 
scheduler and provisioner modules. The algorithms were evaluated on a set 
of scientific workflow ensembles from several different applications using a 
broad range of budget and deadline parameters.

@@

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan2010,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in the scenarios we consider in this paper. Our work can be also
regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In this model, virtual machine (VM) instances are requested on-demand. The VMs are billed by the hour, with partial hours being rounded up. Although there are multiple VM types with different amounts of CPU, memory, disk space, and I/O, for this thesis we focus on a single VM type because we assume that for most applications there will typically be only one or two VM types with the best price/performance ratio for the application~\cite{Juve2009}. We also assume that a submitted task has exclusive access to a VM instance and that there is no preemption.

We assume that there is a delay between the time that a new VM instance is requested and when it becomes available to execute tasks. In practice this delay is typically on the order of tens of seconds to a few minutes, and is highly dependent upon the cloud platform and the VM image size~\cite{Nurmi2008b}.

\subsection{Application Model}

The target applications are ensembles of scientific workflows that can be modeled as Directed Acyclic Graphs (DAGs), where the nodes in the graph represent computational tasks, and the edges represent data- or control-flow dependencies between the tasks.

We assume that there are runtime estimates for each task in the workflow based on either a performance model of the application, or historical data that can be mined. Additionally, we assume that the runtime estimates for individual workflow tasks are not perfect, but may vary based on a uniform distribution of $\pm p\%$.

This study uses synthetic workflows that were generated using historical data from real applications~\cite{Bharathi2008}. The applications come from a wide variety of domains including: bioinformatics (Epigenomics, SIPHT: sRNA identification protocol using high-throughput technology), astronomy (Montage), earthquake science (CyberShake), and physics (LIGO). The runtimes used to generate the synthetic workflows are based on distributions gathered from running real workflows. The synthetic workflows were generated using code developed by Bharathi, et al~\cite{WorkflowGenerator}.

Although scientific workflows are often data-intensive, the algorithms described in the next section do not currently consider the size of input and output data when scheduling tasks. Instead it is assumed that all workflow data is stored in a shared cloud storage system, such as Amazon S3, and that intermediate data transfer times are included in task runtime estimates. It is also assumed that data transfer times between the shared storage and the VMs are equal for different VMs so that task placement decisions do not have an impact on the runtime of the tasks.

This work focuses on scheduling and provisioning for workflow ensembles. A workflow ensemble consists of several related workflows that a user wishes to run. Each workflow in an ensemble is given a numeric priority that indicates how important the workflow is to the user. As such, the priorities indicate the utility function of the user. These priorities are absolute in the sense that completing a workflow with a given priority is more valuable (gives the user more utility) than completing all other workflows in the ensemble with lower priorities combined.

The goal of the workflow ensemble scheduling and cloud provisioning problem is to complete as many high-priority workflows as possible given a fixed budget and deadline. Only workflows for which all tasks are finished by the deadline are considered to be complete---partial results are not usable in this model.

@@

This section describes several algorithms were developed to schedule and provision resources for ensembles of workflows on the cloud under budget and deadline constraints.

\subsection{Static Provisioning Dynamic Scheduling (SPDS) }
\label{sec:spds}

The simplest strategy for executing ensembles of workflows in the cloud is to provision resources statically and schedule workflow tasks dynamically. Given a budget in dollars $b$, deadline in hours $d$, and the hourly price of a VM in dollars $p$, it is easy to calculate the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before the deadline:
%
\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}
%
The SPDS algorithm statically provisions $N_{VM}$ VMs at the start of the ensemble execution and keeps them running until the deadline is reached, or the budget runs out. This provisioning plan has the advantage that it minimizes the number of provisioning and de-provisioning requests.

@@

Initially, the ready tasks from all workflows in the ensemble are added to a priority queue based on the priority of the workflow to which they belong. If there are idle VMs available, and the priority queue is not empty, the next task from the priority queue is submitted to an arbitrarily chosen idle VM. The process is repeated until there are no idle VMs or the priority queue is empty. The scheduler then waits for a task to finish, adds its ready children to the priority queue, marks the VM as idle, and the entire process repeats until the deadline is reached.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/spds-gantt}
  \caption[Example schedule generated by the SPDS algorithm.]{Example schedule generated by the SPDS algorithm. Each row is a different VM. Tasks are represented as boxes and are colored by workflow.}
  \label{fig:spds-example}
\end{figure}

\subsection{Dynamic Provisioning Dynamic Scheduling (DPDS)}

The static provisioning approach of SPDS may perform poorly when resource demand changes during ensemble execution. In those cases, SPDS may either over-provision, wasting the budget on idle VMs, or under-provision, using up the deadline while starving the application of usable VMs. One of the advantages of elasticity in cloud computing is that the number of resources can be changed to meet the demands of an application. In policy-based provisioning (or auto-scaling) systems, metrics such as resource utilization or task queue length are monitored to estimate application demand. This is the approach used by the Dynamic Provisioning Dynamic Scheduling (DPDS) algorithm to determine when to provision and de-provision VMs. The algorithm periodically computes resource utilization using the percentage of idle VMs over time, and adjusts the number of VMs if the utilization is above or below given thresholds. Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are approaching their hourly billing cycle when deciding which VMs to terminate. The DPDS algorithm is shown in Algorithm~\ref{alg:prov}.

@@

The set of VMs completing their billing cycle is determined by considering both the provisioner interval, and the termination delay of the provider. This guarantees that VMs can be terminated before they start the next billing cycle and prevents the budget from being overrun. The VMs terminated in line \ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones that would overrun the budget if not terminated in the current provisioning cycle. The VMs terminated in line \ref{l:terminate2} are chosen to increase the resource utilization to the desired threshold. In order to prevent instances from being terminated too quickly, potentially wasting resources that have already been paid for but could be used later, no more than half of the idle resources are terminated during each provisioning cycle. To avoid an uncontrolled increase in the number of instances, which may happen in the case of highly parallel workflows, the provisioner will not start a new VM if the number of running VMs is greater than the product of $N_{VM}$ (from Equation~\ref{eq:static-plan}) and an autoscaling parameter, $v_{max}$. Unless otherwise specified, $v_{max}$ is assumed to be 1.

@@

The DPDS algorithm does not use any information about the structure of the workflows in the ensemble when scheduling tasks: it looks only at the priorities of the ready tasks when deciding which task to schedule next. It does not consider whether a lower priority task belongs to a workflow that will never be able to complete given the current budget and deadline. As a result, DPDS may start lower priority tasks just to keep VMs busy that will end up delaying higher priority tasks later on, making it less likely that higher priority workflows will be able to finish.

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends DPDS by introducing a workflow admission procedure. The admission procedure is invoked whenever WA-DPDS sees the first task of a new workflow at the head of the priority queue (i.e. when no other tasks from the workflow have been scheduled yet). The admission procedure---shown in Algorithm~\ref{alg:wa-dpds}---estimates whether there is enough budget remaining to admit the new workflow; if there is not, then the workflow is rejected and its tasks are removed from the queue. This algorithm compares the current cost (consumed budget) and remaining budget, taking into account the cost of currently running VMs, and the cost of workflows that have already been admitted. In addition, it adds a small safety margin of \$0.10 to avoid going over the budget.

@@

\subsection{Static Provisioning Static Scheduling (SPSS)}

@@

The approach used by SPSS is to plan each workflow in the ensemble in priority order, rejecting any workflows that cannot be completed by the deadline or that cause the cost of the plan to exceed the budget. Once the plan is complete, the VMs are provisioned and tasks are executed according to the schedules given by the plan.

The disadvantage of the static planning approach used by SPSS is that it is sensitive to dynamic changes in the environment and the application that may disrupt its carefully constructed plan. For example, if there are provisioning delays, or if the runtime estimates for the tasks are inaccurate, then workflow execution may diverge from the plan. This issue will be discussed further in Sections~\ref{sec:delays} and \ref{sec:variances}.

@@

SPSS assigns sub-deadlines to each task based on the slack time of the workflow, which is defined as the amount of extra time that a workflow can extend its critical path and still be completed by the ensemble deadline. For a workflow $w$, the slack time of $w$ is:
%
$$
ST(w) = d - CP(w)
$$
%
where $d$ is the ensemble deadline and $CP(w)$ is the critical path of $w$. We assume that $CP(w) \leq d$, otherwise the workflow cannot be completed by the deadline and must be rejected. For large ensembles we expect the critical path of any individual workflow to be much less than the deadline.

A task's level is the length of the longest path between the task and an entry task of the workflow:
%
$$
Level(t) = \begin{cases}
0,&$if~$Pred(t) = \emptyset\cr
\max_{p \in Pred(t)}Level(p)+1,&$otherwise.$
\end{cases}
$$
%
SPSS distributes the slack time of the workflow by level, so that each level of the workflow gets a portion of the workflow's slack time proportional to the number of tasks in the level and the total runtime of tasks in the level. The idea is that levels containing many tasks and large runtimes should be given a larger portion of the slack time so that tasks in those levels may be serialized. Otherwise, many resources need to be allocated to run all of the tasks in parallel, which may be more costly.

The slack time of a level $l$ in workflow $w$ is given by:
%
$$
ST(l) = ST(w) \Biggl[\left({\alpha}\frac{N(l)}{N(w)}\right) + \left({(1 - \alpha)}\frac{R(l)}{R(w)} \right)\Biggr]
$$
%
where $N(w)$ is the number of tasks in the workflow, $N(l)$ is the number of tasks in level $l$, $R(w)$ is the total runtime of all tasks in the workflow, $R(w)$ is the total runtime of all tasks in level $l$, and $\alpha$ is a parameter between 0 and 1 that causes more slack time to be given to levels with more tasks (large $\alpha$) or more runtime (small $\alpha$).

The deadline of a task $t$ is then:
%
$$
DL(t) = LST(t) + RT(t) + ST(Level(t))
$$
%
where $Level(t)$ is the level of $t$, $RT(t)$ is the runtime of $t$, and $LST(t)$ is the latest start time of $t$ determined by:
%
$$
LST(t) = \begin{cases}
0,&$if~$Pred(t) = \emptyset\cr
\max_{p \in Pred(t)} DL(p),&$otherwise.$
\end{cases}
$$

@@

\begin{figure}[tb] 
  \centering
  \includegraphics[width=\textwidth]{figures/spss-gantt}
  \caption[Example schedule generated by the SPSS algorithm]{Example schedule generated by the SPSS algorithm. Each row is a different VM. Tasks are represented as boxes and are colored by workflow.}
  \label{fig:spss-example}
\end{figure}

\section{Evaluation}
\label{sec:performance}

This section describes a simulation study of the relative performance of three algorithms: DPDS, WA-DPDS, and SPSS. SPDS is not evaluated because, being a simplified version of DPDS and WA-DPDS, it will always produce results worse than the other algorithms.

@@

To evaluate and compare the algorithms, we developed a cloud workflow simulator based on CloudSim~\cite{Calheiros2011}. Our simulation model consists of Cloud, VM and WorkflowEngine entities. The Cloud entity is responsible for staring and terminating VM entities using an API similar to that used by Amazon EC2. VM entities simulate the execution of individual tasks, including randomized variations in runtime. And the WorkflowEngine entity manages the scheduling of tasks and the provisioning of VMs according to the rules established by the different algorithms. We assume that the VMs have a single core and execute tasks sequentially. Although CloudSim provides a more advanced infrastructure model which includes time-sharing and space-sharing policies, we do not use these features since we are interested mainly in the execution of tasks on VMs and high-level workflow scheduling and provisioning. The simulator reads workflow description files in a modified version of the DAX format used by the Pegasus Workflow Management System~\cite{Deelman2005}. The modified file format was created for the synthetic workflow generator developed by Bharathi, et al~\cite{Bharathi2008}.

\subsection{Workflow Ensembles}
\label{sec:ensembles}

In order to evaluate the algorithms on a standard set of workflows, we created randomized ensembles using workflows available from the workflow generator gallery published online by Bharathi, et al~\cite{WorkflowGenerator}. The gallery contains synthetic workflows modeled using structures and parameters that were taken from real applications. Ensembles were created using synthetic workflows from five real applications: SIPHT, a bioinformatics application, LIGO, a physics application, Epigenomics, another bioinformatics application, Montage, an astronomy application, and CyberShake, an earthquake science application. For each of these applications, workflows with 50, 100, 200, 300, 400, 500, 600, 700, 800, 900 and 1000 tasks were generated. For each workflow size, 20 different workflow instances were generated using parameters and task runtime distributions from traces of real workflows. The total collection of synthetic workflows contains 5 applications, 11 different workflow sizes, and 20 workflow instances, for a total of 1100 synthetic workflows.

Using this collection of workflows, we constructed five different ensemble types:

\begin{itemize}
  \item constant
  \item uniform, sorted
  \item uniform, unsorted
  \item Pareto, sorted
  \item Pareto, unsorted
\end{itemize}

In the unsorted ensembles, workflows of different sizes are mixed together, and the priorities are assigned randomly so that there is no relationship between priority and workflow size. For many applications, however, large workflows are more important to users than small workflows because the represent more significant computations. To model this, the sorted ensembles are sorted by size, so that the largest workflows are given the highest priority.

Constant ensembles contain workflows that all have the same number of tasks. For each constant ensemble, the number of tasks is chosen randomly from the set of possible workflow sizes. Once the size is determined, then N workflows of that size are chosen randomly for the ensemble from the set of synthetic workflows for the given application.

Uniform ensembles contain workflows with sizes that are uniformly distributed among the set of possible sizes. Each workflow in a uniform ensemble is selected by first randomly choosing the size the workflow according to a uniform distribution, and then randomly choosing a workflow of that size from the set of synthetic workflows for the given application.

Pareto ensembles contain a small number of larger workflows and a large number of smaller workflows. The sizes of the workflows in a Pareto ensemble are chosen according to a Pareto distribution. The distribution was modified so that the number of large workflows (of size $\geq$ 900) is increased by a small amount to produce a ``heavy-tail''. This causes Pareto ensembles to have a slightly larger number of large workflows, which reflects behavior commonly observed in many computational workloads. An example of the distribution of workflow sizes that occurs in a Pareto ensemble is shown in Figure~\ref{fig:ensemble-pareto}.

\begin{figure}[ht] 
    \centering
    \includegraphics[width=0.6\columnwidth]{figures/ensemble-pareto}
    \caption[Histogram of workflow sizes in Pareto ensembles]{Histogram of workflow sizes in Pareto ensembles. Workflow size is measured in number of tasks.}
    \label{fig:ensemble-pareto}
\end{figure}

The number of workflows in an ensemble depends on the particular application, but we assume that ensemble sizes are on the order of between 10 and 100 workflows. This range is motivated by several reasons: First, such sizes are typical of the real applications we have examined. For example, the number of geographical sites of interest to the users of the CyberShake application in the past has been on the order of 100. Second, smaller ensembles consisting of just a few workflows can be aggregated into a single workflow, so there is no need to treat them as an ensemble. Similarly, when the number of workflows grows, and each workflow has a large number of tasks, either the deadline and budget constraints are low enough to prevent many of the workflows from running, or the problem of efficiently allocating them to the resources becomes similar to a bag-of-tasks problem, which is easier to solve efficiently.

@@

\TODO{Edit this to remove duplication with section \ref{sec:ensembles}}

@@

In order to evaluate the algorithms on a standard set of workflows, we used
ensembles constructed from the workflows available\footnote{\url{https://confluence.pegasus.isi.edu/display/pegasus/WorkflowGenerator}}
in the workflow generator gallery~\cite{Bharathi2008}. The gallery contains synthetic
workflows modeled on real applications. We used workflows from five applications 
that were managed by the Pegasus Workflow Management System. For each
application we constructed ensembles of 100 workflows by sampling collections of
workflows of various sizes (number of tasks) based on a given distribution.
Two distributions are considered:

@@

\subsection{Performance Metric}
\label{sec:perf_metric}
